Diffusion local times and the time until recombination

### 2015-10-21 notes/diffusion-local-times.Rmd

@@

```{r sim_lineage,cache=TRUE}
tanh_drift
nwalks <- 1e4
ntimes <- 1e3
s <- 0.1
tau <- 10
xx <- seq(-15,15,length.out=500)
tt <- seq(0,tau,length.out=ntimes)
rws <- sim_diffusion( tanh_drift, tau, nwalks, ntimes, s=s, xinit=rnorm(nwalks) )
layout( t(1:2), widths=c(3,1) )
matplot(tt, rws[,1:80],type='l', ylab='position', xlab='time', ylim=range(rws) )
legend("topright",lwd=2,legend=c("prob of het","cline","drift"),col=c("black","green","red"))
plot( (-1)*tanh_drift(xx,s=s), xx, col='red', lwd=2, type='l', ylim=range(rws), ylab='', xlab='drift or prob' )
lines( tanh_cline(xx,s=s), xx, col='green', lwd=2 )
lines( (1-s)*(1-tanh_cline(xx,s=s))/(1-s*(1-tanh_cline(xx,s=s))), xx, lwd=2 )
```
The cline is in green and the local probability of being in a heterozygote is in black;
the drift (multiplied by $-1$) is in red.

@@

So, let's solve that, numerically.
We'll use the `ReacTran` library,
which in 1D solves equations of the form
$$\begin{aligned}
   \frac{\partial C}{\partial t} 
   &= 
   \frac{1}{A} \grad \cdot A \left( D \grad C - v C \right) \\
   &= 
   \left( \frac{1}{A} \grad AD - v \right) \cdot \grad C 
   - \frac{1}{A} \grad A \cdot v C 
   + D \Delta C \\
\end{aligned}$$
which is the same as what we have if we let $D=\sigma^2/2$ and $A=p$ and $v=0$.
```{r solve_pde, cache=TRUE}
thegrid <- setup.grid.1D(x.up=-15, x.down=15, N = 1000)
clinefn <- setup.prop.1D(grid=thegrid, func=tanh_cline, s=s)
hetfn <- setup.prop.1D(grid=thegrid, func=tanh_het, s=s)
xx <- seq(-15,15,length.out=500)
pde.fn <- function (t,y,parms,...) {
    tran <- tran.1D(C=y, A=clinefn, D=1/2, dx=thegrid)$dC
    list( tran - r * hetfn$mid * y )
}
yinit <- rep(1.0,thegrid$N)
tt <- seq(0,100,length.out=500)
r <- 1
pde.soln <- ode.1D( y=yinit, times=tt, func=pde.fn, nspec=1 )
image(pde.soln,legend=TRUE,main="prob of no recomb by t", yaxt='n', ylab='space')
axis(2,at=(pretty(thegrid$x.mid)-min(thegrid$x.mid))/diff(range(thegrid$x.mid)),labels=pretty(thegrid$x.mid))
```

It's reassuring the solution is between 0 and 1.
Let's spot check that with the simulation.
Here it is at $t=`r tau`$:
```{r check_pde_soln, cache=TRUE}
check.t <- function (alpha) {
    t.tau <- which.min(abs(tt-alpha*tau))
    probs <- exp( - weighted_time( rws, tanh_het, s=s, use.t=1:floor(nrow(rws)*alpha) ) )
    plot(rws[1,],probs,xlab='initial location', ylab=paste('probability of no recomb by t=',alpha*tau),ylim=c(0,1))
    lines(lowess(rws[1,],probs,f=0.05),col='red',lwd=2)
    lines(thegrid$x.int,pde.soln[t.tau,],col='green',lwd=2)
    legend("topright",lwd=2,col=c("red","green"),legend=c("mean","numerical solution"))
}
check.t(1)
```
and, at other values of $t$:
```{r check_pde_soln_2, cache=TRUE, depends="check_pde_soln"}
layout(t(1:2))
check.t(0.05)
check.t(0.1)
check.t(0.25)
check.t(0.5)
```
**Looks pretty good.**

@@

Of course, we aren't actually trying to estimate the distribution of $T$ itself,
but rather the probabilities of recombination (the Laplace transform).
Here are these probabilities, as a function of $r$
(the second plots are on a log-log scale):
```{r recomb_prob_comparisons, cache=TRUE, depends="sim_general"}
rvals <- exp(seq(log(.001),log(0.5),length.out=400))
precombs <- lapply( lt.comparison, function (lts) {
        sapply( lts, function (lt) sapply( rvals, function (r) { mean((-1)*expm1(-r*lt)) } ) )
    } )
layout(t(1:2))
for (k in seq_along(precombs)) {
    matplot( rvals, precombs[[k]], type='l', col=1:ncol(precombs[[k]]), main=paste("x =",initvals[k]), ylim=range(precombs), ylab="prob of recomb", xlab='r', lty=1 )
    legend("topleft",lty=1,col=1:ncol(precombs[[k]]),legend=colnames(precombs[[k]]))
}
for (k in seq_along(precombs)) {
    matplot( rvals, precombs[[k]], type='l', col=1:ncol(precombs[[k]]), log='xy', main=paste("x =",initvals[k]), ylim=range(precombs), ylab="prob of recomb", xlab='r', lty=1  )
    legend("topleft",lty=1,col=1:ncol(precombs[[k]]),legend=colnames(precombs[[k]]))
}
```

Here's the same thing for a much longer time:
```{r sim_general_longer, fig.height=3*fig.dim, cache=TRUE}
nwalks <- 1e3
ntimes <- 1e4
s <- 0.1
tau <- 100
initvals <- c(0,-3)
lt.comparison <- lapply( initvals, function (xinit) { list(
        tanh=weighted_time( 
               sim_diffusion( tanh_drift,tau,nwalks,ntimes,s=s, xinit=xinit ),
               tanh_het, s=s
           ),
        reflected=weighted_time(
                -abs(sim_diffusion( function(...){0}, tau, nwalks, ntimes, xinit=xinit )),
                function (x,s) { ifelse(abs(x)<1/(2*sqrt(s)), 1/2, 0) }, s=s
            ),
        reflected_tanh=weighted_time(
                -abs(sim_diffusion( function(...){0}, tau, nwalks, ntimes, xinit=xinit )),
                tanh_het, s=s
            ),
        pushed=weighted_time(
                sim_diffusion( function(x,s){ifelse(x>0,-4*sqrt(s),0)}, tau, nwalks, ntimes, s=s, xinit=xinit ),
                function (x) { x>0 }
            ),
        pushed_tanh=weighted_time(
                sim_diffusion( function(x,s){ifelse(x>0,-4*sqrt(s),0)}, tau, nwalks, ntimes, s=s, xinit=xinit ),
                tanh_het, s=s
            )
        ) } )
tbreaks <- seq(0,tau,length.out=50)
Tdists <- lapply( lt.comparison, function (lts) {
        sapply( lts, function (lt) tabulate(findInterval(lt,tbreaks),nbins=length(tbreaks)-1) )
    } )
rvals <- exp(seq(log(.001),log(0.5),length.out=400))
precombs <- lapply( lt.comparison, function (lts) {
        sapply( lts, function (lt) sapply( rvals, function (r) { mean((-1)*expm1(-r*lt)) } ) )
    } )
layout(t(1:2))
for (k in seq_along(precombs)) {
    matplot( tbreaks[-1]-diff(tbreaks)/2, Tdists[[k]], type='l', col=1:ncol(Tdists[[k]]), main=paste("x =",initvals[k]), ylim=range(Tdists), xlab="time in hets", lty=1 )
    legend("topleft",lty=1,col=1:ncol(precombs[[k]]),legend=colnames(precombs[[k]]))
}
for (k in seq_along(precombs)) {
    matplot( rvals, precombs[[k]], type='l', col=1:ncol(precombs[[k]]), main=paste("x =",initvals[k]), ylim=range(precombs), ylab="prob of recomb", xlab='r', lty=1 )
    legend("topleft",lty=1,col=1:ncol(precombs[[k]]),legend=colnames(precombs[[k]]))
}
```

>>>>

The goal
========

The key quantity we compute with diffusion theory 
is the probability that a lineage, 
begun at location $x$ and killed at rate $r$ when in heterozygotes,
is still alive at time $\tau$.
Denote this by
$$
\begin{aligned}
    \zeta(x,r,\tau,\sigma).
\end{aligned}
$$

@@

```{r sim_lineage,cache=TRUE}
tanh_drift
nwalks <- 1e4
ntimes <- 4e3
s <- 0.1
tau <- 5
xx <- seq(-15,15,length.out=500)
tt <- seq(0,tau,length.out=ntimes)
rws <- sim_diffusion( tanh_drift, tau, nwalks, ntimes, s=s, xinit=rnorm(nwalks) )
layout( t(1:2), widths=c(3,1) )
matplot(tt, rws[,1:80],type='l', ylab='position', xlab='time', ylim=range(rws) )
legend("topright",lwd=2,legend=c("prob of het","cline","drift"),col=c("black","green","red"))
plot( (-1)*tanh_drift(xx,s=s), xx, col='red', lwd=2, type='l', ylim=range(rws), ylab='', xlab='drift or prob' )
lines( tanh_cline(xx,s=s), xx, col='green', lwd=2 )
lines( (1-s)*(1-tanh_cline(xx,s=s))/(1-s*(1-tanh_cline(xx,s=s))), xx, lwd=2 )
```
The cline is in green and the local probability of being in a heterozygote is in black;
the drift (multiplied by $-1$) is in red.

@@

So, let's solve that, numerically.
We'll use the `ReacTran` library,
which in 1D solves equations with advection-diffusion parts of the form:
$$\begin{aligned}
   \frac{\partial C}{\partial t} 
   &= 
   \frac{1}{A} \grad \cdot A \left( D \grad C - v C \right) \\
   &= 
   \left( \frac{1}{A} \grad AD - v \right) \cdot \grad C 
   - \frac{1}{A} \grad A \cdot v C 
   + D \Delta C \\
\end{aligned}$$
which is the same as what we have if we let $D=\sigma^2/2$ and $A=p$ and $v=0$.
Here is the setup:
```{r solve_pde_setup, cache=TRUE}
thegrid <- setup.grid.1D(x.up=-15, x.down=15, N = 5000)
clinefn <- setup.prop.1D(grid=thegrid, func=tanh_cline, s=s)
hetfn <- setup.prop.1D(grid=thegrid, func=tanh_het, s=s)
yinit <- rep(1.0,thegrid$N)
tt <- seq(0,tau,length.out=100)
r <- 1
```

First, two quick checks: First, by setting $f_H=1$,
we should have the solution $u(x,t,r) = \exp(-tr)$.
Then, by setting $\sigma$ to be very small, the solution at $x$ should be $\exp(-trf_H(x))$.
```{r solve_pde_check, cache=TRUE, depends="solve_pde_setup"}
layout(t(1:2))
test.hetfn <- setup.prop.1D(grid=thegrid, func=function(x){rep(1,length(x))})
pde.fn <- function (t,y,parms,...) {
    tran <- tran.1D(C=y, A=clinefn, D=1/2, dx=thegrid)$dC
    list( tran - r * test.hetfn$mid * y )
}
pde.soln <- ode.1D( y=yinit, times=tt, func=pde.fn, nspec=1 )
matplot(pde.soln,type='l',ylab="probability of no recomb", col='black',log='y')
lines(tt,exp(-r*tt),lwd=2,lty=2,col='red')
## no diffusion
pde.fn <- function (t,y,parms,...) {
    tran <- tran.1D(C=y, A=clinefn, D=1e-6, dx=thegrid)$dC
    list( tran - r * hetfn$mid * y )
}
pde.soln <- ode.1D( y=yinit, times=tt, func=pde.fn, nspec=1 )
plot.x <- sort(match(sample(thegrid$x.mid[thegrid$x.mid>-5 & thegrid$x.mid<5],12),thegrid$x.mid))
matplot(pde.soln[,plot.x],type='l',ylab="probability of no recomb", col=rainbow(length(plot.x)), lty=1 )
matpoints( exp(outer( tt, -r*hetfn$mid[plot.x], "*" )), lty=4, col=rainbow(length(plot.x)), pch=20 )
```
**Looks good.**

Now, to solve the real problem:
```{r solve_pde, cache=TRUE, depends="solve_pde_setup"}
pde.fn <- function (t,y,parms,...) {
    tran <- tran.1D(C=y, A=clinefn, D=1/2, dx=thegrid)$dC
    list( tran - r * hetfn$mid * y )
}
pde.soln <- ode.1D( y=yinit, times=tt, func=pde.fn, nspec=1 )
image(pde.soln,legend=TRUE,main="prob of no recomb by t", yaxt='n', ylab='space')
axis(2,at=(pretty(thegrid$x.mid)-min(thegrid$x.mid))/diff(range(thegrid$x.mid)),labels=pretty(thegrid$x.mid))
```

It's reassuring the solution is between 0 and 1.
Let's spot check that with the simulation.
Here it is at $t=`r tau`$:
```{r check_pde_soln, cache=TRUE}
check.t <- function (alpha) {
    t.tau <- which.min(abs(tt-alpha*tau))
    probs <- exp( - weighted_time( rws, tanh_het, s=s, use.t=1:floor(nrow(rws)*alpha) ) )
    plot(rws[1,],probs,xlab='initial location', ylab=paste('probability of no recomb by t=',alpha*tau),ylim=c(0,1))
    lines(lowess(rws[1,],probs,f=0.05),col='red',lwd=2)
    lines(thegrid$x.int,pde.soln[t.tau,],col='green',lwd=2)
    legend("topright",lwd=2,col=c("red","green"),legend=c("mean","numerical solution"))
}
check.t(1)
```
and, at other values of $t$:
```{r check_pde_soln_2, cache=TRUE, depends="check_pde_soln"}
layout(t(1:2))
check.t(0.1)
check.t(0.2)
check.t(0.4)
check.t(0.8)
```
**Looks pretty good.**

@@

Of course, we aren't actually trying to estimate the distribution of $T$ itself,
but rather the probabilities of recombination (the Laplace transform).
Here are these probabilities, as a function of $r$
(the second plots are on a log-log scale):
```{r recomb_prob_comparisons, cache=TRUE, depends="sim_general"}
rvals <- exp(seq(log(.001),log(0.5),length.out=400))
precombs <- lapply( lt.comparison, function (lts) {
        sapply( lts, function (lt) sapply( rvals, function (r) { mean((-1)*expm1(-r*lt)) } ) )
    } )
layout(t(1:2))
for (k in seq_along(precombs)) {
    matplot( rvals, precombs[[k]], type='l', col=1:ncol(precombs[[k]]), main=paste("x =",initvals[k]), ylim=range(precombs), ylab="prob of recomb", xlab='r', lty=1 )
    legend("topright",lty=1,col=1:ncol(precombs[[k]]),legend=colnames(precombs[[k]]))
}
for (k in seq_along(precombs)) {
    matplot( rvals, precombs[[k]], type='l', col=1:ncol(precombs[[k]]), log='xy', main=paste("x =",initvals[k]), ylim=range(precombs), ylab="prob of recomb", xlab='r', lty=1  )
    legend("bottomright",lty=1,col=1:ncol(precombs[[k]]),legend=colnames(precombs[[k]]))
}
```

Here's the same thing for a much longer time:
```{r sim_general_longer, fig.height=fig.dim, cache=TRUE}
nwalks <- 1e3
ntimes <- 1e4
s <- 0.1
tau <- 100
initvals <- c(0,-3)
lt.comparison <- lapply( initvals, function (xinit) { list(
        tanh=weighted_time( 
               sim_diffusion( tanh_drift,tau,nwalks,ntimes,s=s, xinit=xinit ),
               tanh_het, s=s
           ),
        reflected=weighted_time(
                -abs(sim_diffusion( function(...){0}, tau, nwalks, ntimes, xinit=xinit )),
                function (x,s) { ifelse(abs(x)<1/(2*sqrt(s)), 1/2, 0) }, s=s
            ),
        reflected_tanh=weighted_time(
                -abs(sim_diffusion( function(...){0}, tau, nwalks, ntimes, xinit=xinit )),
                tanh_het, s=s
            ),
        pushed=weighted_time(
                sim_diffusion( function(x,s){ifelse(x>0,-4*sqrt(s),0)}, tau, nwalks, ntimes, s=s, xinit=xinit ),
                function (x) { x>0 }
            ),
        pushed_tanh=weighted_time(
                sim_diffusion( function(x,s){ifelse(x>0,-4*sqrt(s),0)}, tau, nwalks, ntimes, s=s, xinit=xinit ),
                tanh_het, s=s
            )
        ) } )
tbreaks <- seq(0,tau,length.out=50)
Tdists <- lapply( lt.comparison, function (lts) {
        sapply( lts, function (lt) tabulate(findInterval(lt,tbreaks),nbins=length(tbreaks)-1) )
    } )
rvals <- exp(seq(log(.001),log(0.5),length.out=400))
precombs <- lapply( lt.comparison, function (lts) {
        sapply( lts, function (lt) sapply( rvals, function (r) { mean((-1)*expm1(-r*lt)) } ) )
    } )
layout(t(1:2))
for (k in seq_along(precombs)) {
    matplot( tbreaks[-1]-diff(tbreaks)/2, Tdists[[k]], type='l', col=1:ncol(Tdists[[k]]), main=paste("x =",initvals[k]), ylim=range(Tdists), xlab="time in hets", lty=1 )
    legend("topleft",lty=1,col=1:ncol(precombs[[k]]),legend=colnames(precombs[[k]]))
}
for (k in seq_along(precombs)) {
    matplot( rvals, precombs[[k]], type='l', col=1:ncol(precombs[[k]]), main=paste("x =",initvals[k]), ylim=range(precombs), ylab="prob of recomb", xlab='r', lty=1 )
    legend("topleft",lty=1,col=1:ncol(precombs[[k]]),legend=colnames(precombs[[k]]))
}
```
